Notes and other text

### 2023-04-11 doc/manuscript/translocation.qmd
Draft first three paragraphs of discussion

<!--# Bree comment: It seems to me that this paragraph might be better as a shorter high level compelling summary and then the subsequent sections of the Discussion go into each of the four lines of evidence in more detail. Otherwise it feels a bit heavy and I'm not sure readers will be able to hold it all in mind at once.  -->Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is in fact the mechanism underlying recovery.
In the MYL frog system, there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
First, in host-pathogen systems evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain permanent immunity to Bd and are susceptible to re-infection (Knapp, PeerJ), and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) <!--# This first line of evidence is weak and could be dropped. Just because conditions under which ER is most likely exist in MYL frogs doesn't mean much regarding whether ER is the most likely mechanism underlying recovery. -->.
Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering populations.
These differences are consistent with strong selection and some regions of the genome under selection are associated with known cellular mechanisms associated with resistance, such those related amphibian immune responses <!--# Add specifics to make this stronger, broaden to include not just allele frequencies but genomic regions. -->.
Third, our translocations and long-term assessment of establishment dynamics demonstrate that evolved resistance extends beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frogs and recruits into the adult population in the presence of Bd.
Finally, resistance can be statistically quantified --- in recovering populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics) <!--# Whether this resistance is due to evolution or adaptive immunity is not distinguishable under this line of evidence. -->.
Although other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), few have demonstrated evolutionary rescue: disease-induced selection of host resistance and subsequent host recovery.
Our study is unique in that we provide strong evidence that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.

The ability to harness evolutionary rescue to promote species recovery across a landscape of disease has critical implications for the conservation for MYL frogs.

<!--# The next 2 paragraphs are new text from Allie -->Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs. uninfected *Xenopus tropicalis* [@rosenblum2009].
This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.

@@

-   <!--# Allie text -->One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
    For MYLF evidence suggests that they probably need a bit MORE of an immune response because of findings showing very little response in susceptible animals (Rosenblum et al. 2012) however many studies have found that immune overactivation is linked to susceptibility for Bd.
    This could be different for populations with different histories of Bd exposure.
    Perhaps Rana muscosa has a more recent history with Bd (maybe with another lineage) than for instance Atelopus.
    All of this has led to me think this whole question deserves another paper integrating the Atelopus and MYLF datasets, so maybe we save all this for that?
    Open to whatever you all think is relevant to this story.

@@

-   The MYL frog system is unique in that low productivity due to short summers provide a homogenizing effect across habitat patches, potentially minimizing the effect of the local environment on landscape-level recovery compared to what we might expect in higher productivity systems (for example, tropical amphibian assemblages) <!--# Roland, I am just brainstorming here and this might not make sense. We can delete all of this if necessary. --> .
    That being said, even in this low productivity system we still observed a strong site and environmental effect on translocated adult survival.
    <!--# This is an interesting topic, but despite low productivity MYL frog habitats are highly variable: ponds, streams, meadows, alpine lakes, forested lakes, etc. I think we should drop this bullet.  -->

@@

>>>>

<!--# Bree comment: It seems to me that this paragraph might be better as a shorter high level compelling summary and then the subsequent sections of the Discussion go into each of the four lines of evidence in more detail. Otherwise it feels a bit heavy and I'm not sure readers will be able to hold it all in mind at once. Mark here: I have broken this paragraph up in to two and tried to keep descriptions pretty short.  -->

We provide three lines of evidence that support evolutionary rescue as a primary mechanism of recovery in MYL frog systems.
First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> populations, consistent with strong selection.
Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008a; @ellison2014] (for more information, see SI X).
For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering population and contains eight genes associated with either the MHC Class I or Class III regions.
These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
Broadly, our genomic analyses imply that selection and rapid evolutionary change has contributed to survival in MYL frogs in the presence of Bd.
<!--# For brevity in the discussion, I tried to a pull a few key bullets out of Allie's text. Allie and Bree: Edit and expand as you see fit. -->

The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistance phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
Moreover, resistance has a quantifiable phenotype in this system -- resistant, persisting populations have Bd infection intensities below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
Translocated populations maintain infection intensities below this threshold of infection intensity such infection intensity is not a significant predictor of adult survival following translocations to new environments.
Third and finally, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
Although previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], our study is unique in that these three lines of evidence provide a strong case that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.

Implications for recovery of Mountain yellow legged frogs

@@

-   <!--# Allie text -->

    One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
    For MYLF evidence suggests that they probably need a bit MORE of an immune response because of findings showing very little response in susceptible animals (Rosenblum et al. 2012) however many studies have found that immune overactivation is linked to susceptibility for Bd.
    This could be different for populations with different histories of Bd exposure.
    Perhaps Rana muscosa has a more recent history with Bd (maybe with another lineage) than for instance Atelopus.
    All of this has led to me think this whole question deserves another paper integrating the Atelopus and MYLF datasets, so maybe we save all this for that?
    Open to whatever you all think is relevant to this story.

@@

-   The MYL frog system is unique in that low productivity due to short summers provide a homogenizing effect across habitat patches, potentially minimizing the effect of the local environment on landscape-level recovery compared to what we might expect in higher productivity systems (for example, tropical amphibian assemblages) <!--# Roland, I am just brainstorming here and this might not make sense. We can delete all of this if necessary. --> .
    That being said, even in this low productivity system we still observed a strong site and environmental effect on translocated adult survival.
    <!--# This is an interesting topic, but despite low productivity MYL frog habitats are highly variable: ponds, streams, meadows, alpine lakes, forested lakes, etc. I think we should drop this bullet. Mark here:  -->

@@

<!--# Previous text from Allie below -->

<!--# The next 2 paragraphs are new text from Allie -->

Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs. uninfected *Xenopus tropicalis* [@rosenblum2009].
This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.
